---
title: "Blah blah"
author: "January Weiner"
date: "`r Sys.Date()`"
outputoff: html_document
output:
  xaringan::moon_reader:
    self-contained: true
    css: ["default", "files/cubi-fonts.css", "files/style.css" ]
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
      ratio: "16:9"
toc: no
bibliography: bibliography.bib
nocite: "@weiner2022venn"
---

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=TRUE,autodep=TRUE, results="hide")
```

```{r libraries,cache=FALSE}
library(tidyverse)
library(limma)
library(GEOquery)
library(DESeq2)
## install bioshmods from github.com/bihealth/bioshmods
library(bioshmods)
library(tmod)
library(cowplot)
library(stringr)
library(ggvenn)
library(ggplot2)
library(ggpval)
#library(eulerr)
library(Biobase)
library(pander)
library(org.Hs.eg.db)
library(pwr)
theme_set(theme_minimal() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()))
source("functions.R")
```

```{r setup}
padj_thr <- 0.05
lfc_thr  <- 1
n_stratum <- 10 # 2 strata (male and female) per group/treatment
                # total number of samples = n_stratum * 2 * 2 * 2
lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
```


```{r pheno_data_download,eval=FALSE}
## this sometimes fails. The RDS of the object is provided
Sys.setenv(VROOM_CONNECTION_SIZE=8*131072)

## with getGEO, we can only get the phenoData
geo <- getGEO("GSE156063")[[1]]
saveRDS(geo, file="GSE156063.geo")
```

```{r pheno_data_cleanup}
geo <- readRDS("GSE156063.geo")
covar <- as(phenoData(geo), "data.frame")

## remove columns with only one value
boring <- map_lgl(covar, ~ length(unique(.x)) == 1)
covar <- covar[ , !boring ] 

## clean up
covar <- covar %>% 
  dplyr::rename(gender = "gender:ch1") %>%
  mutate(disease = gsub(" .*", "", .data[["disease state:ch1"]])) %>%
  mutate(label = description) %>%
  mutate(group = disease) %>%
  arrange(description) %>%
  dplyr::select(all_of(c("title", "label", "gender", "disease", "group"))) 
rownames(covar) <- covar$label
```

```{r featuredata_download}
## the counts must be downloaded from GEO separately.
if(!file.exists("GSE156063_swab_gene_counts.csv.gz")) {
  download.file("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE156nnn/GSE156063/suppl/GSE156063_swab_gene_counts.csv.gz",
                "GSE156063_swab_gene_counts.csv.gz")
}

counts <- read_csv("GSE156063_swab_gene_counts.csv.gz")
.tmp <- counts[[1]]
counts <- as.matrix(counts[,-1])
rownames(counts) <- .tmp
#counts <- counts[ , covar$description ]
counts <- counts[ , covar$label ]
lcpm   <- edgeR::cpm(counts, log=TRUE)
#stopifnot(all(colnames(counts) == covar$description))
stopifnot(all(colnames(counts) == covar$label))

annot <- data.frame(ENSEMBL = rownames(counts))

.tmp <- mapIds(org.Hs.eg.db, annot$ENSEMBL, column=c("ENTREZID"), keytype="ENSEMBL")
annot$ENTREZID <- .tmp[ match(annot$ENSEMBL, names(.tmp)) ]

.tmp <- mapIds(org.Hs.eg.db, annot$ENSEMBL, column=c("SYMBOL"), keytype="ENSEMBL")
annot$SYMBOL <- .tmp[ match(annot$ENSEMBL, names(.tmp)) ]

.tmp <- mapIds(org.Hs.eg.db, annot$ENSEMBL, column=c("GENENAME"), keytype="ENSEMBL")
annot$GENENAME <- .tmp[ match(annot$ENSEMBL, names(.tmp)) ]
```


```{r Preparation_of_the_covariates,cache=FALSE}
sel <- covar$group %in% c("no", "SC2")
counts <- counts[ , sel ]
covar  <- covar[ sel, ]
lcpm   <- lcpm[ , sel ]
covar$group <- as.character(covar$group)

set.seed(0123)

#covar$disease <- covar$group

g1 <- covar %>% mutate(n=1:n()) %>%
  group_by(gender, disease) %>% slice_sample(n=n_stratum) %>% pull(n)
g2 <- covar %>% mutate(n=1:n()) %>% filter(!n %in% g1) %>%
  group_by(gender, disease) %>% slice_sample(n=n_stratum) %>% pull(n)

covar$group <- NA
covar$group[g1] <- "G1"
covar$group[g2] <- "G2"

covar$group.disease <- paste0(covar$group, '_', covar$disease)

```

```{r DESeq2,cache=FALSE}
## DESeq2 calculations
## manual cache, since the operation takes a long time
sel <- !is.na(covar$group)

ds2 <- DESeqDataSetFromMatrix(counts[,sel], colData=covar[sel, ], design=~ 0 + group.disease )

ds2_file <- "ds2_cached.rds"
if(!file.exists(ds2_file)) {
  message("Running DESeq2")
  ds2 <- DESeq(ds2)
  saveRDS(ds2, file=ds2_file)
} else {
  message("Reading ds2 from manual cache")
  ds2 <- readRDS(ds2_file)
}
```

```{r DE_analysis}
library(tidyverse)
library(tmod)
res <- list()
res$g1 <- results(ds2, contrast=c(-1, 1, 0, 0))
res$g2 <- results(ds2, contrast=c(0, 0, -1, 1))

res <- map(res, ~ .x %>% as.data.frame() %>%
  rownames_to_column("ENSEMBL") %>% 
  left_join(annot, by="ENSEMBL") %>%
  mutate(DEG=!is.na(padj) & abs(log2FoldChange) > lfc_thr & padj < padj_thr)) 

res.merged <- merge(res$g1, res$g2, by=c("ENSEMBL", "SYMBOL", "ENTREZID", "GENENAME"), suffixes=c(".g1", ".g2"))
```



```{r GO_db_prepare,cache=TRUE}
library(msigdbr)
min_mod_size <- 10
max_mod_size <- 50
msig_go_bp <- msigdbr(subcategory="GO:BP")
mset <- makeTmodFromDataFrame(msig_go_bp, 
                              feature_col="human_gene_symbol", 
                              module_col="gs_exact_source", 
                              title_col="gs_name")
mset$gs$N <- map_int(mset$gs2gv, length)
mset <- mset[ mset$gs$N <= max_mod_size & 
              mset$gs$N >= min_mod_size ]
```


```{r GSEA,cache=TRUE}
common <- res.merged %>% filter(DEG.g1 & DEG.g2) %>% pull(SYMBOL)
gsea_res <- map(res, ~ {
  fg <- .x %>% filter(padj < padj_thr & abs(log2FoldChange) > lfc_thr)
  fg <- setdiff(fg$SYMBOL, common) # only "specific" genes
  tmodHGtest(fg=fg, bg=.x$SYMBOL, mset=mset)
})
gsea_res_full <- map(res, ~ {
  fg <- .x %>% filter(padj < padj_thr & abs(log2FoldChange) > lfc_thr)
  fg <- setdiff(fg$SYMBOL, common) # only "specific" genes
  tmodHGtest(fg=fg, bg=.x$SYMBOL, mset=mset, qval = Inf)
})
message("G1 ", nrow(gsea_res$g1), " G2 ", nrow(gsea_res$g2))
gsea_res.merged <- merge(gsea_res$g1, gsea_res$g2, by=c("ID", "Title"), suffixes=c(".g1", ".g2"), all=T)
saveRDS(gsea_res.merged, file="gsea_res_merged.rds")
```


```{r lit_surveys}
## collect the result of the literature surveys
get_keywords <- function(file, pattern="^([a-zA-Z]+): *(.*)") {
  ll <- readLines(file)
  ll <- ll[ grepl(pattern, ll) & !grepl("^https://", ll) ]

  ret <- gsub(pattern, "\\2", ll)
  names(ret) <- gsub(pattern, "\\1", ll)
  ret <- as.list(ret)
  ret$file <- file
  ret$incorrect_count <- get_incorrect_counts(file)
  ret$incorrect_count_vdonly <- get_incorrect_counts(file, vdonly=TRUE)
  ret
}

get_incorrect_counts <- function(file, vdonly=FALSE) {

  ll <- readLines(file)

  pat <- "^https://"
  if(vdonly) {
    sum(grepl(pat, ll) & grepl("vd only", ll))
  } else {
    sum(grepl(pat, ll))
  }

}

files <- list.files(pattern="literature_survey.*.md")

finfo <- map(files, get_keywords) 
names(finfo) <- map_chr(finfo, ~ .x[["journal"]])
```





## Gene set enrichments in two groups of patients

```{r fig1, fig.width=12,fig.height=4.5,dpi=600,dev="png"}
p1 <- ggvenn(list(G1=res$g1 %>% filter(DEG) %>% pull(ENSEMBL),
                         G2=res$g2 %>% filter(DEG) %>% pull(ENSEMBL)),
             show_percentage=FALSE) +
      labs(title="Differentially expressed genes") +
      ylim(-1, 1.4)
p2 <- ggvenn(list(G1=gsea_res$g1 %>% pull(ID),
                         G2=gsea_res$g2 %>% pull(ID)),
              show_percentage=FALSE) +
      labs(title="Enriched GO terms") +
      ylim(-1, 1.4)

col1 <- plot_grid(p1, p2, labels="AUTO", nrow=2)

p3 <- plot_gsea(gsea_res$g1 %>% arrange(P.Value) %>% dplyr::slice(1:10)) + ggtitle("G1") +
  scale_x_discrete(labels=function(x) str_wrap(x, width=25))
#p3 <- plot_grid(p3, NULL, ncol=1, rel_heights=c(2, 1))
p4 <- plot_gsea(gsea_res$g2 %>% arrange(P.Value) %>% dplyr::slice(1:10)) + ggtitle("G2") +
  scale_x_discrete(labels=function(x) str_wrap(x, width=25))

plot_grid(plotlist = list(col1, p3, p4), labels=c('', 'C', 'D'), nrow=1)
```

"The Difference Between “Significant” and “Not Significant” is not
Itself Statistically Significant" – Andrew Gelman and Hal Stern

---

##

.pull-left[

 * In 
 * Not the only problem with GSE!



]

.pull-right[

<div syle="font-size:small;">
Weiner 3rd, January, Benedikt Obermayer, and Dieter Beule. "Venn diagrams may indicate erroneous statistical reasoning in transcriptomics." Frontiers in Genetics 13 (2022).
</div>

]

